# Reproducible Research: Peer Assessment 1

``` {r setoptions, echo=FALSE} 
# set gobal options
opts_chunk$set(echo=TRUE)
```

## Loading and preprocessing the data
Set working directory 
```{r workingdirectory, echo=TRUE}
# set working directory
setwd("./courseradatascience/RepData_PeerAssessment1")
```

### Loading data  
```{r loadata}
# assign subdirectory where data resides
localZipFile <- "activity.zip"

# assign local zip file 
subdir <- "activity"

# if necesary unzip the data file into activity directory
if (!file.exists(subdir)) {
   unzip(localZipFile)
}

# read data
data <- read.csv("./activity/activity.csv", header=TRUE)

```

### Tranforming data to a format suitable for analysis  
```{r transformingdata}
# convert date string to Date class 
data$date <- as.Date(data$date)
```

## What is mean total number of steps taken per day?

### Histogram of the total number of steps takey each day 
```{r averages}
library(plyr)
# aggregate the number of steps taken each day 
df0 <- ddply(data, "date", summarise, total=sum(steps, na.rm=TRUE))

# present a summary of total number of steps per day  
summary(df0)

# draw histogram. assume 1000 steps bin interval from summary above 
library(ggplot2)
ggplot(df0, aes(x=total)) + 
    geom_histogram(binwidth=1000) + 
    scale_y_continuous(limits=c(0,25)) + 
    labs(title="Histogram of Total Number of Steps Taken per Day Oct/Nov 2012", x="Total Number of Steps Daily", y="Frequency")
```

### Calculate mean and median of total number of steps per day 
```{r calculatemeans}
steps.mean <- formatC(round(mean(df0$total, na.rm=TRUE)), format="d", big.mark=",")
steps.median <- formatC(round(median(df0$total, na.rm=TRUE)), format="d", big.mark=",")
```
The mean of total number of steps taken each day is `r steps.mean`    
The median of total number of steps taken each day is `r steps.median`   
   
## What is the average daily activity pattern?
### Time series plot of 5-minute interval(x-axis) over average number of steps taken, averaged across all days (y-axis) 
```{r dailyaverages}
# average number of steps taken on each 5-min interval across all days 
df1 <- ddply(data, "interval", summarise, average=mean(steps, na.rm=TRUE))

# plot 5-min intervals (x-axis) and average number of steps taken across all-days (y-axis)
ggplot(df1, aes(x=interval, y=average)) + 
    geom_line() + 
    labs(title="Daily Activity Pattern Oct/Nov 2012", x="5-Minutes Interval", y="Average Number of Steps Taken")
```
### 5-minute interval, on average across all days in the dataset, containing the maximum number of steps
```{r maximumnumberofsteps}
z <- round(max(df1$average))
max.interval <- df1[df1$average>z, ][1,1]
```
The 5-minute interval number (on average across all days) containing the maximum number of steps is `r max.interval`  

## Inputing missing values
### Calculating and reporting the number of missing values in the dataset
```{r reportingmissingvalues}
# calculate number of NA values 
missing.steps <- formatC(sum(is.na(data$steps)>0), format="d", big.mark=",")
missing.date <- formatC(sum(is.na(data$date)>0), format="d", big.mark=",")
missing.interval <- formatC(sum(is.na(data$interval)>0), format="d", big.mark=",")
```
Number of missing step values is `r missing.steps`   
Number of missing date values is `r missing.date`   
Number of missing interval values is `r missing.interval`    

### Strategy handling replacing missing values: replace missing values with mean for corresponding 5-minute interval
```{r replacingmissingvalues}
# recalculate means for 5-minute interval (rounded)  
df2 <- ddply(data, "interval", summarise, average=round(mean(steps, na.rm=TRUE)))

# identify rows with missing values 
data.na <- data[is.na(data$steps)>0,]
```

### Creating a new dataset with missing values replaced 
```{r createnewdata}
# copy dataset 
data.new <- data

# replaced values counter 
j <- 0

# loop through new dataset and replace NAs 5-minute interval mean number of steps
for (i in seq(1, nrow(data.new))) {
    # check for missing values
    if(complete.cases(data.new[i,])==FALSE) {
        # identify interval
        interval <- data.new[i, "interval"]
        # replace missing values with average steps taken on corresponding interval 
        data.new[i, "steps"] <- df2[df2$interval==interval, "average"]
        j <- j + 1
    }
}

```
Number of missing values replaced `r j`

### Histogram of total number of steps taken each day with new dataset 
```{r replacementimpact}
library(plyr)
# aggregate the number of steps taken each day 
df3 <- ddply(data.new, "date", summarise, total=sum(steps))

# present a summary of total number of steps per day  
summary(df3)

# draw histogram. assume 1000 steps bin interval from summary above 
library(ggplot2)
ggplot(df3, aes(x=total)) + 
    geom_histogram(binwidth=1000) + 
    scale_y_continuous(limits=c(0,25)) + 
    labs(title="Histogram of Total Number of Steps Taken per Day Oct/Nov 2012\nAfter Missing Values Replacement", x="Total Number of Steps Daily", y="Frequency")
```

### Calculate mean and median of total number of steps per day for the new dataset with imputted missing data  
```{r dailyaveragesafterreplacement}
steps.mean.new <- formatC(round(mean(df3$total, na.rm=TRUE)), format="d", big.mark=",")
steps.median.new <- formatC(round(median(df3$total, na.rm=TRUE)), format="d", big.mark=",")
```
For the new dataset (after missing values replacement):   
The mean of total number of steps taken each day for the new dataset is `r steps.mean.new`    
The median of total number of steps taken each day is `r steps.median.new` 

### Impact of imputting missing data on the estimates of total number of steps  
Original dataset (with missing values removed): Mean = `r steps.mean` - Median = `r steps.median`     
New dataset (after missing values replacement): Mean = `r steps.mean.new` - Median = `r steps.median.new`         
The missing data imputting strategy resulted in a shift of mean and median values to the right in the order of 400 more steps    

## Are there differences in activity patterns between weekdays and weekends?

### Creating a variable to differenciate weekdays from weekend days 
```{r identifyweekdays}
# create factor variable day with two values "weekend" and "weekday"
data.new$day <- as.factor(ifelse(substr(weekdays(data.new$date), 1, 1)=="S", "weekend", "weekday"))
```

### Making a panel plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis) 
```{r panelplot}
# calculate means for 5-minute interval (rounded)  
df4 <- ddply(data.new, c("day", "interval"), summarise, average=round(mean(steps)))

# panel plot for 5-min intervals (x-axis) and average number of steps taken across all-days (y-axis)
ggplot(df4, aes(x=interval, y=average)) + 
    geom_line() +    
    facet_grid(day ~ ., as.table=FALSE) +     
    theme(strip.text.y = element_text(size=12, face="bold")) +     
    labs(title="Daily Activity Pattern Oct/Nov 2012", x="5-Minutes Interval", y="Average Number of Steps Taken")
```